---
title: "Practical_Week7"
author: "pippin"
date: "2025-11-18"
output: html_document
---

## **7.4** Introduction

### **7.4.1** Analysing Spatial Autocorrelation with Moran’s I, LISA and friends

In this section we are going to explore patterns of spatially referenced continuous observations using various measures of spatial autocorrelation. Spatial autocorrelation is a measure of similarity between nearby data. Check out the various references in the reading list for more information about the methods we will explore today. There are also useful links in the help file of the `spdep` package which we will be using here.

This week my question might be **Are the values (in this case the density of blue plaques) similar (or dissimilar) across the wards of London**

### **7.4.2** Data download

Before we get any further, let’s get some ward boundaries read in to R — download LondonWardData from [the London Data store](https://data.london.gov.uk/dataset/statistical-gis-boundary-files-london) and read it in…

**Use the data in the statistical-gis-boundaries-london.zip**

```{r}
library(here)
library(janitor)
library(sf)
library(tidyverse)
#read the ward data in
LondonWards <- st_read(here::here("Week7/Data/statistical-gis-boundaries-london/ESRI", "London_Ward.shp"))
```

```{r}
LondonWardsMerged <- st_read(here::here("Week7/Data/statistical-gis-boundaries-london/ESRI","London_Ward_CityMerged.shp"))%>%
  st_transform(.,27700)
```

```{r}
WardData <- read_csv("https://data.london.gov.uk/download/f33fb38c-cb37-48e3-8298-84c0d3cc5a6c/772d2d64-e8c6-46cb-86f9-e52b4c7851bc/ward-profiles-excel-version.csv",
                    locale = locale(encoding = "latin1"),
                    na = c("NA", "n/a"))%>%
  clean_names()

LondonWardsMerged <- LondonWardsMerged%>%
  left_join(WardData,
            by = c("GSS_CODE" = "new_code"))%>%
  distinct(GSS_CODE, .keep_all = TRUE)%>%
  dplyr::select(GSS_CODE, ward_name, average_gcse_capped_point_scores_2014)
```

It’s probably projected correctly, but in case it isn’t give it a projection using the `st_crs()` function in the `sf` package

```{r}
st_crs(LondonWardsMerged)
```

```{r}
BluePlaques <- st_read("https://s3.eu-west-2.amazonaws.com/openplaques/open-plaques-london-2018-04-08.geojson")%>%
  st_transform(.,27700)

library(tmap)

tmap_mode("plot")
tm_shape(LondonWardsMerged) +
  tm_polygons(fill_alpha = 0.5) +
tm_shape(BluePlaques) +
  tm_dots(fill = "blue", size=0.1)
```

```{r}
summary(BluePlaques)
```

## **7.5** Data cleaning

Ah yes, we might need to lose the blue plaques that fall outside of London

```{r}
BluePlaquesSub <- BluePlaques[LondonWardsMerged,]

tmap_mode("plot")
tm_shape(LondonWardsMerged) +
  tm_polygons(fill_alpha = 0.5) +
tm_shape(BluePlaquesSub) +
  tm_dots(fill = "blue", size=0.1)
```

## **7.6** Data manipulation

The measures of spatial autocorrelation that we will be using require continuous observations (counts of blue plaques, average GCSE scores, average incomes etc.) to be spatially referenced (i.e. attached to a spatial unit like a ward or a borough). The file you have already has the various obervations associated with the London Ward data file already attached to it, but let’s continue with our blue plaques example for now.

To create a continuous observation from the blue plaques data we need to count all of the blue plaques that fall within each Ward in the City. Luckily, we can do this using the `st_intersects()` function from the `sf` package.

`st_intersects()` will return a list of which points (plaques) are in each polygon (wards). So each row will be a polygon and the ID are the point IDs that intersect the polygon.

In this case we don’t want to use `st_join()` (which by default uses `st_intersects()`) as `st_join()` is like a left join and will retain all the data in on the left. So if a ward is empty, there will still be a row for that ward. So, we use `st_intersects()` on its own. We could also use other topological relationship functions such as `st_within()` instead…

```{r}
example<-st_intersects(LondonWardsMerged, BluePlaquesSub)

example
```

the output of st_intersects() is **"sgbp" "list"**

```{r}
class(example)
```

Here the polygon with the ID of 6 Kingston upon Thames - Coombe Hill has three plaques within it…we can check this with st_join (or using QGIS by opening the data)…we have three plaques…But note the ID column added is different to the ID of the data…open blueplaques from the environment window and you will see the IDs that were returned in `st_intersects()`.

```{r}
### just check...
example2<-st_join(LondonWardsMerged, BluePlaquesSub)
```

Now we just take the length of each list per polygon and add this as new column…

```{r}
points_sf_joined <- LondonWardsMerged%>%
  mutate(n = lengths(st_intersects(., BluePlaquesSub)))%>%
  janitor::clean_names()%>%
  #calculate area
  mutate(area=st_area(.))%>%
  #then density of the points per ward
  mutate(density=n/area)%>%
  #select density and some other variables 
  dplyr::select(density, ward_name, gss_code, n, average_gcse_capped_point_scores_2014)
```

How about a quick choropleth map to see how we are getting on…

```{r}
points_sf_joined<- points_sf_joined %>%                    
  group_by(gss_code) %>%         
  summarise(density = first(density),
          wardname= first(ward_name),
          plaquecount= first(n))

tm_shape(points_sf_joined) +
    tm_polygons(
      fill = "density",
      fill.scale = tm_scale_intervals(
        values = "brewer.blues",
        style="jenks"),
      # set the legend
      fill.legend = tm_legend(title="Blue Plaque Density",
                              title.size=0.85,
                              size=0.8,
                              # plot outside of the main map
                              #explained below
                              position=tm_pos_out("right", 
                                                  "center",
                                                  pos.v = "center")))
```

So, from the map, it looks as though we might have some clustering of blue plaques in the centre of London so let’s check this with Moran’s I and some other statistics.

## **7.7** Weight matrix

Before being able to calculate Moran’s I and any similar statistics, we need to first define a Wij spatial weights matrix

```{r}
library(spdep)
```

First calculate the centroids of all Wards in London

```{r}
coordsW <- points_sf_joined%>%
  st_centroid()%>%
  st_geometry()
  
plot(coordsW,axes=TRUE)
```

Now we need to generate a spatial weights matrix (remember from the lecture). We’ll start with a simple binary matrix of queen’s case neighbours (otherwise known as Contiguity edges corners). This method means that polygons with a shared edge or a corner will be included in computations for the target polygon…A **spatial weight matrix represents the spatial element of our data**, this means we are trying to conceptualize and model how parts of the data are linked (or not linked) to each other spatially, using rules that we will set.

**A note on spatial weight matrices**

If the features share a boundary they are **contiguous**, this can also be classed as only common boundaries — a **rook** (like chess a rook can move forwards or side wards) or any point in common (e.g. corners / other boundaries) — a **queen** (like chess a queen can move forwards, backwards or on a diagonal).

**Alternatively** instead of using contiguous relationships you can use **distance** based relationships. This is frequently done with k nearest neighbours in which k is set to the closest observations. e.g. K=3 means the three closest observations.

In the first instance we must create a neighbours list — which is a list of all the neighbours. To do so we will use with the argument saying we want a to use Queens case. Let’s see a summary of the output`poly2nb()queen=T`

```{r}
#create a neighbours list
LWard_nb <- points_sf_joined %>%
  poly2nb(., queen=T)
```

The output is a neighbour list object hence why i’ve called the object . We can get some summary info and plot the matrix.`LWard_nb`

```{r}
summary(LWard_nb)
```

Here it is telling us that the average number of neighbours is 5.88. Now let’s plot the neighbours.

```{r}
#plot them
plot(LWard_nb, st_geometry(coordsW), col="red")
#add a map underneath
plot(points_sf_joined$geometry, add=T)
```

### **7.7.1** Matrix style

From the weights list we must now make a spatial weight matrix. This makes a matrix the size of the number of neighbourhoods with values indicating if the elements in the rows are a neighbour or not. The style of weight is especially important here:

-   B is the basic binary coding (1/0)

-   W is **row** standardised (sums over all links to n)

-   C is **globally** standardised (sums over all links to n)

-   U is equal to C divided by the number of neighbours (sums over all links to unity)

-   S is the variance-stabilizing coding scheme proposed by Tiefelsdorf et al. 1999, p. 167-168 (sums over all links to n).

To understand this further let’s start with binary.

```{r}
#create a spatial weights matrix from these weights
Lward.lw <- LWard_nb %>%
  nb2mat(., style="B")

sum(Lward.lw)
```

Summing the binary (1/0) shows that we have 3680 neighbours. Now if we were to do **global** standardisation this would mean dividing our 625 wards by the total number of neighbours meaning each spatial weight has a value of 0.169.

Alternatively we can do **row** standardisation where 1 is divided by the sum of the number of neighbours in each row. For example, row 1 here sums to 6, meaning each weight would be 0.166 in row 1 only.

```{r}
sum(Lward.lw[1,])
```

Standardisation permits comparable spatial parameters. But you also need to consider geographical features — should a neighbour be a neighbour if there is a river between them?

## **7.8** Autocorrelation

Now we have defined our Wij matrix, we can calculate the Moran’s I and other associated statistics. However, Moran’s I requires a spatial weight **list** type object as opposed to **matrix**, this is simply…

```{r}
Lward.lw <- LWard_nb %>%
  nb2listw(., style="C")
```

### **7.8.1** Moran’s I

Moran’s I test tells us whether we have clustered values (close to 1) or dispersed values (close to -1). We will calculate for densities rather than raw values (think about why this is?).

Here:

-   Clustered refers to how similar values are to the neighbours. This could be *any value* - high or low.

-   Dispersed means how dissimilar values are to neighbours

-   Random values would produce a result of 0.

Our neighbours are set by the weight matrix.

```{r}
I_LWard_Global_Density <- points_sf_joined %>%
  pull(density) %>%
  as.vector()%>%
  moran.test(., Lward.lw)

I_LWard_Global_Density
```

### **7.8.2** Geary’s C

Geary’s C as well..? This tells us whether **similar values or dissimilar values are clustering**. Here:

-   C \< 1 indicates that neighbours with similar values are clustered (compared to random chance)

-   C \>1 indicates that neighbours with dissimilar values are clustered (compared to random chance).

-   C = 0 indicates not autocorrelation

```{r}
C_LWard_Global_Density <- 
  points_sf_joined %>%
  pull(density) %>%
  as.vector()%>%
  geary.test(., Lward.lw)

C_LWard_Global_Density
```

#### **7.8.2.1** Comparison

We should now have an idea about what the main differences between Moran’s I and Geary’s C are, and how to inptert the outputs. But how do these look in an equation. I really **dislike using equations** as a descritpiton is much easier to understand. However, it’s important to see how similar these two autocorrelcation measures are.

#### **7.8.2.2** Key message

For Moran’s I

-   we are looking at the product of deviaitions from the global mean (subtracting our central value (zi−¯z) and niehgbour (zj−¯z) from the global mean individually, then multiplying them).

-   If both **deviations are positive then this suggests clustering**

For Geary’s C

-   This changes to the difference between the central value and the neighbour, squared (zi−zj)2

-   A **large difference means more dissimiarlity between the values**

### **7.8.3** Getis Ord

Getis Ord General G…? This tells goes further and says if **high or low values are clustering**. Here:

-   If G \> expected = High values clustering, which could be termed a “hot spot”

-   If G \< expected = low values clustering, which could termed a “cold spot”

-   If G = expected, then it is spatially random

```{r}
G_LWard_Global_Density <- 
  points_sf_joined %>%
  pull(density) %>%
  as.vector()%>%
  globalG.test(., Lward.lw)

G_LWard_Global_Density
```

**Note** conceptaully Gertis Ord General G is again very similar to Moran’s I and Geary’s C but the focus is now on the *weighted product of values* - this means multiplying neighbours and weighting them with our weight matrix (in the numerator). **If a high value is near a high value it returns a larger product**.

### **7.8.4** Summary

The Moran’s I statistic = 0.67 (remember 1 = clustered, 0 = no pattern, -1 = dispersed) which shows that we have some distinctive clustering. But this could be clustering of any value.

-   Interpretation: We have some clustering of values (low or high) *across the entire study area*.

The Geary’s C statistic = 0.41 (remember Geary’s C falls between 0 and 2; 1 means no spatial autocorrelation, \<1 - positive spatial autocorrelation or similar values clustering, \>1 - negative spatial autocorreation or dissimilar values clustering), which shows that similar values are clustering

-   Interpretation: We have some clustering of similar values *indicating neighbours are similar*.

The General G statistic = G \> expected, so high values are tending to cluster.

-   Interpretation: We have clustering *of high values*.

### **7.8.5** Local Moran’s I

We can now also calculate local versions of the Moran’s I statistic (e.g. for each Ward) and a Getis Ord G∗i statistic to spatially see and map *where* we have clusters…

This iteraties over each individual polygon (or point), using just the neighbours for the spatial unit from our weight matrix.

Local Moran’s I is:

-   The difference between a value and the global mean (zi−¯z) \* the difference between a neighbour and the global mean wij(zj−¯z), weighted by the weight matrix

-   This is standarised by the global variance (m2) to make sure values are comparable.

Ii=(zi−¯z)m2∑jwij(zj−¯z)

**Where:**

-   Ii: Local Moran’s I statistic for unit *i*

-   zi: Value at location *i*

-   zj: Value at neighboring location *j*

-   ¯z: Mean of all (global) values

-   wij: Spatial weight between locations *i* and *j*

-   m2=1n∑nk=1(zk−¯z)2: Global variance

    -   zk: is each individual value in the dataset

It returns several columns, of most interest is the Z score. A Z-score is **how many standard deviations a value is away (above or below) from the mean**. This allows us to state if our value is significantly different than expected value at this location considering the neighours.

We are comparing our value of Moran’s I to that of an expected value (computed from a separate equation that uses the spatial weight matrix, and therefore considers the neighbouring values).

We are expecting our value of Moran’s I to be in the middle of the distribution of the expected values. These expected values follow a normal distribution, with the middle part representing complete spatial randomness. This is **typically** between \< -1.65 or \> +1.65 standard deviations from the mean

The **null hypothesis** is always there is complete spatial randomness. A null hypothesis means:

> no statistical significance exists in a set of given observations

If our value is towards the tails of the distribution then it is unlikely that the value is completely spatially random and we can reject the null hypothesis…as it is not what we expect at this location.

In the example where we use a z-score of \>2.58 or \<-2.58 we interpret this as…

> …\> 2.58 or \<-2.58 standard deviations away from the mean are significant at the 99% level…this means there is a \<1% chance that autocorrelation is not present, indicating **positive autocorrelation** (this could be **high surrounded by high or low surrounded by low**).

A low or negative Z score is indicative of spatial outlier (e.g. high next to low values)

The [Global vs location spatial autocorrelation resource](https://storymaps.arcgis.com/stories/5b26f25bb81a437b89003423505e2f71) goes through the specific formulas here, but the most important parts are knowing

-   What we are comparing values to in Local Moran’s I

-   What the results mean

-   Why the results could be important

```{r}
#use the localmoran function to generate I for each ward in the city

I_LWard_Local_count <- points_sf_joined %>%
  pull(plaquecount) %>%
  as.vector()%>%
  localmoran(., Lward.lw)%>%
  as_tibble()

I_LWard_Local_Density <- points_sf_joined %>%
  pull(density) %>%
  as.vector()%>%
  localmoran(., Lward.lw)%>%
  as_tibble()

#what does the output (the localMoran object) look like?
slice_head(I_LWard_Local_Density, n=20)
```

There are 20 columns of data. We want to copy some of the columns (the I score (column 1) and the z-score standard deviation (column 4)) back into the LondonWards spatialPolygonsDataframe

```{r}
points_sf_joined <- points_sf_joined %>%
  mutate(plaque_count_I = as.numeric(I_LWard_Local_count$Ii))%>%
  mutate(plaque_count_Iz =as.numeric(I_LWard_Local_count$Z.Ii))%>%
  mutate(density_I =as.numeric(I_LWard_Local_Density$Ii))%>%
  mutate(density_Iz =as.numeric(I_LWard_Local_Density$Z.Ii))
```

#### **7.8.5.1** Mapping

No we can plot a map of the local Moran’s I outputs…

We’ll set the breaks manually based on the rule that data points \>2.58 or \<-2.58 standard deviations away from the mean are significant at the 99% level (\<1% chance that autocorrelation not present); \>1.96 - \<2.58 or \<-1.96 to \>-2.58 standard deviations are significant at the 95% level (\<5% change that autocorrelation not present). \>1.65 = 90% etc.

```{r}
breaks1<-c(-1000,-2.58,-1.96,-1.65,1.65,1.96,2.58,1000)
```

```{r}
library(RColorBrewer)
MoranColours<- rev(brewer.pal(7, "RdGy"))
```

Plot the output

<https://andrewmaclachlan.github.io/CASA0005repo/spatial-autocorrelation.html>

HERE—reverse colour scale??

```{r}
tm_shape(points_sf_joined) +
    tm_polygons(
      fill = "plaque_count_Iz",
      fill.scale = tm_scale_intervals(values = "brewer.br_bg",
                                      midpoint = "NA",
                                      style = "fixed",
                                      breaks=breaks1),
       fill.legend = tm_legend(title="Local Moran's I, Blue Plaques in London",
                              title.size=0.85,
                              size=0.8,
                              # plot outside of the main map
                              #explained below
                              position=tm_pos_out("right", 
                                                  "center",
                                                  pos.v = "center")))
```

```{r}
I_LWard_Local_Density%>%
  colnames()
```

This map shows some areas in the centre of London that have relatively high scores, indicating areas with lots of blue plaques neighbouring other areas with lots of blue plaques.

### **7.8.6** Local Getis Ord G∗i

What about the Getis Ord G∗i statistic for hot and cold spots…

This is a very similar concept to Local Moran’s I except it just returns a z-score…remember that a z-score shows **how many standard deviations a value (our value) is away (above or below) from the mean (of the expected values)**

In Moran’s I this is just the value of the spatial unit (e.g. polygon of the ward) compared to the neighbouring units.

Here, to be significant (or a hot spot) we will have a high value surrounded by high values. The local sum of these values will be different to the expected sum (think of this as all the values in the area) then where this difference is large we can consider it to be not by chance…

The same z-score criteria then applies as before..

This summary from L3 Harris nicely summaries the Getis Ord G∗i output…

> The result of Getis Ord G∗i analysis is an array of Z-scores, one for each pixel [or polygon], which is the number of standard deviations that the pixel [or polygon] and its neighbours are from the global mean. High Z-scores indicate more intense clustering of high pixel values, indicating hot spots. Low Z-scores indicate more intense clustering of low values, indicating cold spots. Individual pixels with high or low values by themselves might be interesting but not necessarily significant.

```{r}
Gi_LWard_Local_Density <- points_sf_joined %>%
  pull(density) %>%
  as.vector()%>%
  localG(., Lward.lw)

head(Gi_LWard_Local_Density)
```

```{r}
points_sf_joined <- points_sf_joined %>%
  mutate(density_G = as.numeric(Gi_LWard_Local_Density))
```

```{r}
tm_shape(points_sf_joined) +
    tm_polygons(
      fill = "density_G",
      fill.scale = tm_scale_intervals(values = "brewer.br_bg",
                                      midpoint = "NA",
                                      style = "fixed",
                                      breaks=breaks1),
       fill.legend = tm_legend(title="Gi*, Blue Plaques in London",
                              title.size=0.85,
                              size=0.8,
                              # plot outside of the main map
                              #explained below
                              position=tm_pos_out("right", 
                                                  "center",
                                                  pos.v = "center")))
```

### Summary

| Statistic | What is compared | Purpose |
|------------------|------------------|-----------------------------------|
| **Local Moran’s I** | Deviation of a point from neibhouring values | Identify clusters or spatial outliers (similar/dissimilar) |
| **Moran’s I Z-score** | Observed Moran’s I vs expected under randomness | Measure statistical significance of spatial autocorrelation at each location; high values surrounded by high values or low values surrounded by low values |
| **Getis-Ord** $G_i^*$ | Local weighted sum vs global sum of all values | Identify hotspots (high values clustering) or coldspots (low values clustering) |

## 7.9 Other variables

The local Moran’s I and G ∗ i statistics for wards clearly show that the density of blue plaques in the centre of the city exhibits strong (and positive) spatial autocorrelation, but neither of these maps are very interesting. Why not try some alternative variables and see what patterns emerge… here I’m going to have a look at Average GSCE scores…

```{r}
#use head to see what other variables are in the data file

slice_head(points_sf_joined, n=2)
```

```{r}
Datatypelist <- LondonWardsMerged %>% 
  st_drop_geometry()%>%
  summarise_all(class) %>%
  pivot_longer(everything(), 
               names_to="All_variables", 
               values_to="Variable_class")

Datatypelist
```

```{r}
I_LWard_Local_GCSE <- LondonWardsMerged %>%
  arrange(GSS_CODE)%>%
  pull(average_gcse_capped_point_scores_2014) %>%
  as.vector()%>%
  localmoran(., Lward.lw)%>%
  as_tibble()

points_sf_joined <- points_sf_joined %>%
  arrange(gss_code)%>%
  mutate(GCSE_LocIz = as.numeric(I_LWard_Local_GCSE$Z.Ii))

tm_shape(points_sf_joined) +
    tm_polygons(
      fill = "GCSE_LocIz",
      fill.scale = tm_scale_intervals(values = "brewer.br_bg",
                                      midpoint = "NA",
                                      style = "fixed",
                                      breaks=breaks1),
       fill.legend = tm_legend(title="Local Moran's I, GCSE Scores",
                              title.size=0.85,
                              size=0.8,
                              # plot outside of the main map
                              #explained below
                              position=tm_pos_out("right", 
                                                  "center",
                                                  pos.v = "center")))
```

Now the Gi\* statistic to look at clusters of high and low scores and explain what the output map is showing and what other questions this can lead us to ask next week…..

```{r}
G_LWard_Local_GCSE <- LondonWardsMerged %>%
  dplyr::arrange(GSS_CODE)%>%
  dplyr::pull(average_gcse_capped_point_scores_2014) %>%
  as.vector()%>%
  localG(., Lward.lw)

points_sf_joined <- points_sf_joined %>%
  dplyr::arrange(gss_code)%>%
  dplyr::mutate(GCSE_LocGiz = as.numeric(G_LWard_Local_GCSE))


tm_shape(points_sf_joined) +
    tm_polygons(
      fill = "GCSE_LocGiz",
      fill.scale = tm_scale_intervals(values = "brewer.br_bg",
                                      midpoint = "NA",
                                      style = "fixed",
                                      breaks=breaks1),
       fill.legend = tm_legend(title="Gi*, GCSE Scores",
                              title.size=0.85,
                              size=0.8,
                              # plot outside of the main map
                              #explained below
                              position=tm_pos_out("right", 
                                                  "center",
                                                  pos.v = "center")))
```
